A practical review on linear and nonlinear global approaches to flow instabilities

### 2018-12-09 99_Documentation/ARTICLE_STABFEM/ARTICLE_ASME_Revision.tex
final (?) version of the paper

%%% first author
\author{D. Fabre$^{(a)}$,  V. Citro$^{(b,a)}$,  D. Ferreira Sabino$^{(a)}$, P. Bonnefis$^{(a)}$,  F. Giannetti$^{(b)}$ \& M. Pigou$^{(a)}$ 
    \affiliation{
	$^{(a)}$ Institut de M\'ecanique des fluides de Toulouse (IMFT), University of Toulouse \\
	$^{(b)}$ Dipartimento di Ingegneria (DIIN), Universit\'a} di Salerno. 
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
{\it 
This paper aims at reviewing linear and nonlinear approaches to study the stability of fluid flows. 
We provide a concise but self-contained exposition of the main concepts and specific numerical methods 
designed for global stability studies, including the classical linear stability analysis, the adjoint-based sensitivity and  
the most recent nonlinear developments. 
Regarding numerical implementation, a number of ideas making resolution particularly efficient are discussed, 
including mesh adaptation, simple shift-invert strategy instead of the classical Arnoldi algorithm, 
and a simplification of the recent nonlinear self-consistent approach proposed by Manti\v{c}-Lugo et al. (2014). 
An open-source software implementing all the concepts discussed in the present paper is provided. 
The software is demonstrated for the reference case of the incompressible, two-dimensional flow around a circular cylinder, but is easily customisable to a variety of other flow configurations or flow equations.
%An abstract for an ASME paper should be less than 150 words and is normally in italics.
}
\end{abstract}

@@

The mesh generated by the previous way may not be optimal for the stability calculations as the structure of the eigenmode may be more complex than that of the base flow. 
To tackle this aspect, the idea is to subsequently adapt the mesh to both the mesh flow and the results of the stability calculation.
This is easily done with FreeFem, as the  AdaptMesh procedure can be used with several objective functions.
We have experimented four different strategies.
The first (D strategy) is to adapt the mesh to the base flow and the structure of the leading direct eigenmode.
A second strategy (A) is to use the adjoint mode instead of the direct mode for mesh adaptation.
However, as the sensitivity of the eigenvalue to perturbations of the operator (including discretization errors) is more closely linked to structural sensitivity concepts, it sounds a better idea to use either the 'wavemaker' $S_w$ or the endogeneity $S_e$ to adapt the mesh, leading to the two last experimented strategies, called S and E.
Note that using the S strategy was already experimented in {\em \cite{GiannettiLuchini} for the wake of a cylinder (IS THAT THE RIGHT REFERENCE ??}).
Mesh adaptation using the scalar product of the direct and adjoint eigenmodes (which turns out to be equivalent to the E strategy introduced here considering the definition of the endogeneity) was also proposed in \cite{mavripilis2015adjoint}.
A detailed comparison of the four strategies is reported in appendix A.
It is shown that for the case of a cylinder, all strategies give the same values for base-flow drag, eigenvalues and several properties of the nonlinear limit cycle (see sec. 4) with less than $0.3\%$ deviation, but that the A, E and especially the S strategy lead to significantly lower number of grid points compared to the D strategy.
The practical conclusions of this study are the following recommendations:

\begin{itemize}
\item[-] 
First, if one is only interested in the eigenvalues --for instance to plot growth rates as function of Reynolds or to identify the critical Reynolds number,-- the A, E or S strategies are the best options. In the case of the cylinder, the S strategy leads to the lightest mesh. The E strategy results in a more refined mesh as the function $S_e$ is complex and displays more gradients than the real-valued function $S_w$ (as illustrated in figure \ref{fig:StructEndo}); hence the sensitivity revealed by the phase of $S_e$, which may be significant, is not captured by using the simpler function $S_w$. In the case of the flow around a cylinder,  this loss of information has no practical effect on the results, but for usage with more complex flows, our recommendation for future studies will be to favour the E strategy.
\item[-] 
The situation is different if one wants to design a mesh allowing to study the flow structure in the whole domain (for instance for plotting its structure). In that case, using A,E or S strategy will allow to accurately compute the eigenvalue and the structure of the mode in the wavemaker region, but will not allow for a proper resolution in the regions located far away where the direct mode may reach large amplitude. This is particularly the case for open flows characterized by large spatial convective amplifications. In such cases the D strategy is the only way to get a proper description of the eigenmode in the whole domain.
\end{itemize}

@@

The main concepts of global linear stability analysis, exposed in section 2 for the simplest case of incompressible flows, are directly generalizable to more complex situations involving compressibility, fluid-structure interactions, deformable free surfaces, etc... Such more complex situations can also be studied using the same class of numerical methods and the same methodology. One of the ambitions of the StabFem software if to provide a unified procedure to perform such stability studies, so that despite the fact that the underlying equations may be different, the corresponding Octave/Matlab scripts will be almost the same in all cases. In this section, we will illustrate this point for the case of a compressible flow, keeping the geometry as a 2D cylinder, and will reproduce the results of the recent study of \cite{Fani2018}.

@@

In our implementation, these calculations are done by the FreeFem++ solvers {\sf  Newton2DComp.edp} and {\sf  Stab2DComp.edp}, which are quite different from their incompressible counterparts, but these solvers are wrapped by the same generic drivers {\sf  SF_BaseFlow.m}  and {\sf  SF_Stability.m}. The selection of which solver to use is made according to the parameters transmitted to them.

@@

 \begin{figure}
\includegraphics[width=\linewidth]{Cylinder_EigenmodeRe60Ma02_vort.png}
\caption{Structure of the unstable eigenmode (streamwise velocity component) for the compressible flow around a cylinder  $(Re= 150,M=0.2)$.}
\label{fig:Eigenmode}
\end{figure}

As an illustration, we consider the compressible flow around a 2D cylinder (same geometry as in previous section). The reader may find on the website of the project a script \footnote{ \url{https://gitlab.com/stabfem/StabFem/blob/master/STABLE_CASES/compCYLINDER/SRIPT_Fani.m}} reproducing the main results of the study of \cite{Fani2018}. We restrict here to the illustration of the structure of the unstable eigenmode for $Re = 150; M = 0.2$. Figure \ref{fig:compcyl_ux} shows the axial velocity associated to the eigenmode in a domain centered on the body. Figure \ref{fig:compcyl_p} displays the stucture of the pressure field on a much larger domain, allowing to detail the structure of the radiated acoustic field. The structure is in perfect agreement with the results of \cite{Fani2018} (see 5$(a)$ and $8$ of this paper). The value of the corresponding eigenvalue is also perfectly reproduced, namely $\lambda = 0.446+0.697i$.

 \begin{figure}
\includegraphics[width=\linewidth]{Cylinder_EigenmodeRe60Ma02_p.png}
\caption{Structure of the unstable eigenmode (pressure component) for the compressible flow around a cylinder  $(Re= 150,M=0.2)$.}
\label{fig:Eigenmode}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{Cylinder_Strouhal_Re_HB.png}
\includegraphics[width=\linewidth]{Cylinder_Cx_Re_HB.png}
\includegraphics[width=\linewidth]{Cylinder_Cy_Re_SC.png}
\includegraphics[width=\linewidth]{Cylinder_Energy_Re_SC_AdaptD.png}
\includegraphics[width=\linewidth]{Cylinder_Lx_Re_HB.png}
\end{center}
\caption{Comparison between the weakly nonlinear results (WNL), the harmonic balance data (HB1), and baseflow/linear results: Strouhal number $(a)$, Mean drag $(b)$, amplitude of oscillating lift $(c)$, Energy-amplitude of the nonlinear perturbation $(d)$ and recirculation length of mean/base flows $(e)$. 
%c_x should be consistent with the drag definition given at beginning
}
\label{fig:HB_SC_DATA_COMP}
\end{figure}

@@

The objective of this paper is twofold. First, we aimed at giving an up-to-date and comprehensive review on global stability approaches, both linear and nonlinear, including the most recent developments of the field. 
Secondly, we intended to provide an easy to use software performing all these computations from a single program.
In accordance with this objective, all the figures presented in the paper can be produced by launching a single Octave/Matlab program available on the website of the project.
This program handles all computation steps, from calls to FreeFem++ to figure generation.

Although the focus here was on the reference case of 2D incompressible flow around a cylinder, the {\em StabFem } software is designed to be easily customisable to a variety of other situations.
The project is currently in constant development, and incorporates a growing number of other configurations.
Under its present status, the project incorporates test-cases for the following classes of problems:
\begin{itemize}
\item[-] Incompressible flows around 2D objects, either fixed or in movement such as the case of a spring-mounted cylinder \cite{Navrose},
\item[-] Incompressible flows in axisymmetric geometries such as the wake of spheres and disks \cite{Tchoufag2015} and the flow through apertures \cite{FabreISMA}, 
\item[-] Compressible flow around 2D objects  \cite{Fani2018}, 
\item[-] Oscillations of hanging drops and liquid bridges \cite{Chireux2015}
\end{itemize}
The project is intended as collaborative, so anyone who wants to contribute is welcome !

>>>>

%%% first author
\author{D. Fabre$^{(a)}$,  V. Citro$^{(b,a)}$,  D. Ferreira Sabino$^{(a)}$, P. Bonnefis$^{(a)}$,  J. Sierra $^{(a)}$, F. Giannetti$^{(b)}$ \& M. Pigou$^{(a)}$ 
    \affiliation{
	$^{(a)}$ Institut de M\'ecanique des fluides de Toulouse (IMFT), University of Toulouse \\
	$^{(b)}$ Dipartimento di Ingegneria (DIIN), Universit\'a} di Salerno. 
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
{\it 
This paper aims at reviewing linear and nonlinear approaches to study the stability of fluid flows. 
We provide a concise but self-contained exposition of the main concepts and specific numerical methods 
designed for global stability studies, including the classical linear stability analysis, the adjoint-based sensitivity and  
the most recent nonlinear developments. 
Regarding numerical implementation, a number of ideas making resolution particularly efficient are discussed, 
including mesh adaptation, simple shift-invert strategy instead of the classical Arnoldi algorithm, 
and a simplification of the recent nonlinear self-consistent approach proposed by Manti\v{c}-Lugo et al. (2014). 
An open-source software implementing all the concepts discussed in the present paper is provided. 
The software is demonstrated for the reference case of the two-dimensional flow around a circular cylinder, in both incompressible and compressible cases, but is easily customisable to a variety of other flow configurations or flow equations.
%An abstract for an ASME paper should be less than 150 words and is normally in italics.
}
\end{abstract}

@@

%The mesh generated by the previous way may not be optimal for the stability calculations as the structure of the eigenmode may be more complex than that of the base flow.  To tackle this aspect, the idea is to subsequently adapt the mesh to both the mesh flow and the results of the stability calculation.This is easily done with FreeFem, as the  AdaptMesh procedure can be used with several objective functions.We have experimented four different strategies.The first (D strategy) is to adapt the mesh to the base flow and the structure of the leading direct eigenmode. A second strategy (A) is to use the adjoint mode instead of the direct mode for mesh adaptation. However, as the sensitivity of the eigenvalue to perturbations of the operator (including discretization errors) is more closely linked to structural sensitivity concepts, it sounds a better idea to use either the 'wavemaker' $S_w$ or the endogeneity $S_e$ to adapt the mesh, leading to the two last experimented strategies, called S and E.Note that using the S strategy was already experimented in {\em \cite{GiannettiLuchini} for the wake of a cylinder (IS THAT THE RIGHT REFERENCE ??}).Mesh adaptation using the scalar product of the direct and adjoint eigenmodes (which turns out to be equivalent to the E strategy introduced here considering the definition of the endogeneity) was also proposed in \cite{mavripilis2015adjoint}.A detailed comparison of the four strategies is reported in appendix A.It is shown that for the case of a cylinder, all strategies give the same values for base-flow drag, eigenvalues and several properties of the nonlinear limit cycle (see sec. 4) with less than $0.3\%$ deviation, but that the A, E and especially the S strategy lead to significantly lower number of grid points compared to the D strategy.
%The practical conclusions of this study are the following recommendations:

%\begin{itemize}
%\item[-] 
%First, if one is only interested in the eigenvalues --for instance to plot growth rates as function of Reynolds or to identify the critical Reynolds number,-- the A, E or S strategies are the best options. In the case of the cylinder, the S strategy leads to the lightest mesh. The E strategy results in a more refined mesh as the function $S_e$ is complex and displays more gradients than the real-valued function $S_w$ (as illustrated in figure \ref{fig:StructEndo}); hence the sensitivity revealed by the phase of $S_e$, which may be significant, is not captured by using the simpler function $S_w$. In the case of the flow around a cylinder,  this loss of information has no practical effect on the results, but for usage with more complex flows, our recommendation for future studies will be to favour the E strategy.
%\item[-] 
%The situation is different if one wants to design a mesh allowing to study the flow structure in the whole domain (for instance for plotting its structure). In that case, using A,E or S strategy will allow to accurately compute the eigenvalue and the structure of the mode in the wavemaker region, but will not allow for a proper resolution in the regions located far away where the direct mode may reach large amplitude. This is particularly the case for open flows characterized by large spatial convective amplifications. In such cases the D strategy is the only way to get a proper description of the eigenmode in the whole domain.
%\end{itemize}

The mesh generated in the previous way may not be optimal for the stability calculations as the structure of the eigenmode may be more complex than that of the base flow. To remedy with this, the idea is to subsequently adapt the mesh to both the mesh flow and the results of the stability calculation. This is easily done with FreeFem, as the  AdaptMesh procedure can be used with several objective functions. We have experimented four different strategies. The first (D strategy) is to adapt the mesh to the base flow and the structure of the leading direct eigenmode. A second strategy (A) is to use the adjoint mode instead of the direct mode for mesh adaptation. However, as the sensitivity of the eigenvalue to perturbations of the operator (including discretization errors) is more closely linked to structural sensitivity concepts, it sounds a better idea to use the 'wavemaker' $S_w$ to adapt the mesh, leading to the last experimented strategy, called S. 
Mesh adaptation using the scalar product of the direct and adjoint eigenmodes  (called E-strategy) was also proposed in \cite{mavripilis2015adjoint} (this quantity has also beed discussed by \cite{marquet2015endogeneity}). A detailed comparison of the four strategies is reported in appendix A. It is shown that for the case of a cylinder, all strategies give the same values for base-flow drag, eigenvalues and several properties of the nonlinear limit cycle (see sec. 4) with less than $0,3\%$ deviation, but that the A, E and especially the S strategy lead to significantly lower number of grid points compared to the D strategy. 
Note that a valid alternative (not yet available in the StabFem  project) is offered by the Error Sensitivity to Refinement recently proposed by \cite{Luchini2017}.

@@

The main concepts of global linear stability analysis, exposed in section 2 for the simplest case of incompressible flows, are directly generalizable to more complex situations involving compressibility, fluid-structure interactions, deformable free surfaces, etc... Such more complex situations can be studied using the same class of numerical methods and the same methodology. One of the ambitions of the StabFem software if to provide a unified procedure to perform such stability studies, so that despite the fact that the underlying equations may be different, the corresponding Octave/Matlab scripts will be almost the same in all cases. In this section, we will illustrate this point for the case of a compressible flow, keeping the geometry as a 2D cylinder, and will reproduce the results of the recent study of \cite{Fani2018}.

@@

In our implementation, these calculations are done by the FreeFem++ solvers {\sf  Newton2DComp.edp} and {\sf  Stab2DComp.edp}, which are quite different from their incompressible counterparts. However these solvers are wrapped by the same generic drivers {\sf  SF\_BaseFlow.m}  and {\sf  SF\_Stability.m}. The selection of which solver to use is made according to the parameters transmitted to them.

@@

 \begin{figure}
\includegraphics[width=\linewidth]{Cylinder_EigenmodeRe60Ma02_vort.png}
\caption{Structure of the unstable eigenmode (streamwise velocity component) for the compressible flow around a cylinder  $(Re= 150,M=0.2)$.}
\label{fig:compcyl_ux} 
\end{figure}

As an illustration, we consider the compressible flow around a 2D cylinder (same geometry as in previous section). The reader may find on the website of the project a script \footnote{ \url{https://gitlab.com/stabfem/StabFem/blob/master/STABLE_CASES/compCYLINDER/SRIPT_Fani.m}} reproducing the main results of the study of \cite{Fani2018}. We restrict here to the illustration of the structure of the unstable eigenmode for $Re = 150; M = 0.2$. Figure \ref{fig:compcyl_ux} shows the axial velocity associated to the eigenmode in a domain centered on the body. Figure \ref{fig:compcyl_p} displays the stucture of the pressure field on a much larger domain, allowing to detail the structure of the radiated acoustic field. The structure is in perfect agreement with the results of \cite{Fani2018} (see 5$(a)$ and $8$ of this paper). The value of the corresponding eigenvalue is also perfectly reproduced, namely $\lambda = 0.15273+0.64482i$.

 \begin{figure}
\includegraphics[width=\linewidth]{Cylinder_EigenmodeRe60Ma02_p.png}
\caption{Structure of the unstable eigenmode (pressure component) for the compressible flow around a cylinder  $(Re= 150,M=0.2)$. Upper half : real part ; lower half : imaginary part. }
\label{fig:compcyl_p}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.9\linewidth]{Cylinder_Strouhal_Re_HB.png}
\includegraphics[width=0.9\linewidth]{Cylinder_Cx_Re_HB.png}
\includegraphics[width=0.9\linewidth]{Cylinder_Cy_Re_SC.png}
\includegraphics[width=0.9\linewidth]{Cylinder_Energy_Re_SC_AdaptD.png}
\includegraphics[width=0.9\linewidth]{Cylinder_Lx_Re_HB.png}
\end{center}
\caption{Comparison between the weakly nonlinear results (WNL), the harmonic balance data (HB1), and baseflow/linear results: Strouhal number $(a)$, Mean drag $(b)$, amplitude of oscillating lift $(c)$, Energy-amplitude of the nonlinear perturbation $(d)$ and recirculation length of mean/base flows $(e)$. 
%c_x should be consistent with the drag definition given at beginning
}
\label{fig:HB_SC_DATA_COMP}
\end{figure}

@@

The objective of this paper is twofold. First, we aimed at giving an up-to-date and comprehensive review on global stability approaches, both linear and nonlinear, including the most recent developments of the field. 
Secondly, we intended to provide an easy to use software performing all these computations from a single program.
In accordance with this objective, all the figures presented in the paper can be produced by launching a single Octave/Matlab program available on the website of th project.
This program handles all computation steps, from calls to FreeFem++ to figure generation.

Although the focus here was on the reference case of 2D incompressible and compressible flow around a cylinder, the {\em StabFem } software is designed to be easily customisable to a variety of other situations.
The project is currently in constant development, and incorporates a growing number of other configurations.
Under its present status, the project incorporates test-cases for the following classes of problems:
\begin{itemize}
\item[-] Incompressible flows around 2D and axisymmetric blunt bodies of various geometries, including spheres and disks \cite{Tchoufag2015}. 
\item[-] Incompressible flows around 2D and 3D objects in free movement, including for instance the spring-mounted cylinder \cite{Navrose}, and freely falling disks and spheres \cite{tchoufag2014global}.
\item[-] Incompressible and compressible flow flow through apertures \cite{FabreISMA,longobardi2018IUTAM}, 
\item[-] Compressible flow around 3D objects  \cite{meliga2010effect},
\item[-] Oscillations of hanging drops, sessile drops and liquid bridges \cite{Chireux2015},
\item[-] Rotating free surface flows \cite{mougel2017instabilities}.
\end{itemize}
Our ambition is to use the website of the project as a support to publish scripts allowing anyone to reproduce the main results of our past and future publications
on such topics.  Such scripts are already available concerning Refs. \cite{Tchoufag2015,longobardi2018IUTAM,Chireux2015,mougel2017instabilities} and the list of available case will be growing in the coming months.The project is intended as collaborative, so anyone who wants to contribute is welcome !
